---
title: "CalCofi Database Query"
author: "Marina Kochuten"
format: html
message: false
warning: false
embed-resources: true
theme: lux
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r}
# Load libraries
library(tidyverse)
library(DBI)
library(duckdb)
library(ggthemes)
```

You may need to update the path to the database to match your file structure:

```{r}
# Connect to database
db_path <- '../sql/cal-cofi.db' # Path to database
conn <- DBI::dbConnect(duckdb::duckdb(), dbdir=db_path)
```

## Queries

How does sea temperature and salinity vary over time?

```{r}
# Sea temp averages
avg_temp <- dbGetQuery(conn, "SELECT Year, AVG(T_degC) AS Avg_Temp 
                                FROM (SELECT Cst_Cnt, T_degC, Year 
                                      FROM bottle_table 
                                      JOIN cast_table 
                                      USING (Cst_Cnt)) 
                                      GROUP BY Year")

# Salinity averages
avg_sal <- dbGetQuery(conn, "SELECT Year, AVG(Salnty) AS Avg_Salinity 
                                FROM (SELECT Cst_Cnt, Salnty, Year 
                                      FROM bottle_table 
                                      JOIN cast_table USING (Cst_Cnt)) 
                                      GROUP BY Year")
```

## Plot time series

```{r}
#| fig-width: 9

# Exclude outlier to better visualize overall trend
filtered_temp <- avg_temp |> 
  filter(Avg_Temp < 14)

# Temperature plot
ggplot(filtered_temp, aes(x = Year, y = Avg_Temp)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(y = "Average Sea Temperature (C)",
       title = "Average Annual Sea Temperature From 1950-2020") +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  scale_y_continuous(breaks = scales::breaks_pretty()) +
  theme_wsj() +
  theme(plot.title = element_text(size = 18, hjust = 1, margin = margin(0,0,15,0)),
        axis.title.y = element_text(size = 12,
                                    margin = margin(5, 15, 5, 0)),
        plot.margin = margin(15,20,15,20))
```

This plot shows that since 1950, average sea temperature has been increasing in California.

```{r}
#| fig-width: 9

# Exclude outliers to better visualize overall trend
filtered_sal <- avg_sal |>
  filter(Year != 2021, Avg_Salinity < 35.0)

# Salinity plot
ggplot(filtered_sal, aes(x = Year, y = Avg_Salinity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(y = "Average Annual Salinity (PSS)",
       title = "Average Annual Salinity From 1950 - 2020") +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  scale_y_continuous(breaks = scales::breaks_pretty()) +
  theme_wsj() +
  theme(plot.title = element_text(size = 18, hjust = 1, margin = margin(0,0,15,0)),
        axis.title.y = element_text(size = 12,
                                    margin = margin(5, 15, 5, 0)),
        plot.margin = margin(15,20,15,20))
```

This plot shows that the average ocean salinity in California has been declining since 1950. This phenomena has been observed near the poles as a result of glacial melt introducing freshwater to the oceans, but I find it interesting that this is being observed off the coast of California as well.

```{r}
# Disconnect from database
DBI::dbDisconnect(conn, shutdown = TRUE)
```
